---
title: "Exploratory Data Analysis"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---


```{r include = FALSE}
library(tidyverse)
library(haven)
library(viridis)
library(RColorBrewer)
library(scales)
library(plotly)
library(patchwork)
library(gridExtra)
```


```{r}
ukb_data <- read_dta("./external_data/data_prepared.dta")
```

### Age and Sex Distribution
```{r}
figure_1 <- ukb_data %>%
  mutate(age_group = cut(age, breaks = seq(10, 100, by = 10), right = FALSE, 
                         labels = paste(seq(10, 90, by = 10), "-", 
                                        seq(19, 99, by = 10), sep = ""))) %>%
  ggplot(aes(x = age_group, y = age, fill = as.factor(sex))) + 
  geom_boxplot(alpha = 0.9) + 
  theme_minimal() +
  labs(x = 'Age Groups', 
       y = 'Age',
       title = 'Age Distribution by Sex and Age Groups',
       fill = 'Sex') +
  scale_fill_manual(values = c("lightblue", "lightpink"),
                    labels = c("Female", "Male")) +
  theme(legend.position = "bottom")

figure_1
```

We examined the age distribution of male and female participants across 10-year age groups. In the 30-39 age group, the sample size was relatively small, with median ages and IQRs consistent across sexes. Similar trends were observed in the 40-49 and 50-59 groups. In the 60-69 group, females exhibited slightly higher upper quartiles than males, while the 70-79 group showed broader distributions for females, indicating more variation in ages.

<br>

### Alcohol Consumption and Disease Incidence
```{r message = FALSE}
# Data cleaning
disease_heat2 <- ukb_data %>%
  filter(!is.na(sex), !is.na(total_alcohol), !is.na(nafld_outcome)) %>%
  mutate(
    alcohol_group = cut(total_alcohol, 
                       breaks = c(-Inf, 0, 5, 10, 15, Inf),
                       labels = c("None", "Low", "Medium", "High", "Very High"),
                       include.lowest = TRUE)
  ) %>%
  group_by(sex, alcohol_group) %>%
  summarise(
    disease_rate = mean(nafld_outcome, na.rm = TRUE),
    n = n()
  ) %>%
  ungroup()

ukb_data <- ukb_data %>%
  mutate(
    sex = as_factor(sex), 
    nafld_outcome = as.numeric(as_factor(nafld_outcome)),
    total_alcohol = as.numeric(total_alcohol),
    income = as.numeric(income)
  )

disease_heat2 <- ukb_data %>%
  filter(!is.na(sex), !is.na(total_alcohol), !is.na(nafld_outcome)) %>%
  mutate(
    alcohol_group = cut(total_alcohol, 
                       breaks = c(-Inf, 0, 5, 10, 15, Inf),
                       labels = c("None", "Low", "Medium", "High", "Very High"),
                       include.lowest = TRUE)
  ) %>%
  group_by(sex, alcohol_group) %>%
  summarise(
    disease_rate = mean(nafld_outcome, na.rm = TRUE),
    n = n()
  ) %>%
  ungroup()

# Create heatmap
figure_2 <- ggplot(disease_heat2, aes(x = sex, y = alcohol_group, fill = disease_rate)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  theme_minimal() +
  labs(title = "Disease Incidence by Sex and Alcohol Consumption",
       x = "Sex",
       y = "Alcohol Consumption Level",
       fill = "Disease Rate") +
  geom_text(aes(label = sprintf("n=%d", n)), color = "black", size = 3)

ggplotly(figure_2)
```

A heatmap analysis revealed an inverse relationship between alcohol consumption and disease incidence. Non-drinkers exhibited the highest disease rates, with males outnumbering females. In contrast, participants with very high alcohol consumption had the lowest disease rates. Across all categories, females demonstrated slightly higher disease rates than males.

<br>

### Blood Pressure and Age Trends
```{r message = FALSE}
# Data Cleaning
disease_heat1 <- ukb_data %>%
  filter(!is.na(age), !is.na(bp_cat), !is.na(nafld_outcome)) %>%
  mutate(
    bp_cat = as_factor(bp_cat),
    age_group = cut(age, 
                   breaks = c(30, 40, 50, 60, 70, 80),
                   labels = c("30-40", "40-50", "50-60", "60-70", "70-80"),
                   include.lowest = TRUE)
  ) %>%
  group_by(age_group, bp_cat) %>%
  summarise(
    disease_rate = mean(nafld_outcome, na.rm = TRUE),
    n = n()
  ) %>%
  ungroup()

# Create heatmap 
figure_4 <- ggplot(disease_heat1, aes(x = age_group, y = bp_cat, fill = disease_rate)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  labs(title = "Disease Incidence by Age and Blood Pressure Category",
       x = "Age Groups",
       y = "BP Category",
       fill = "Disease Rate") +
  geom_text(aes(label = sprintf("n=%d", n)), color = "black", size = 3)

ggplotly(figure_4)
```

We observed that disease incidence increased significantly with higher blood pressure, particularly in individuals aged 60 and above. Younger age groups (30-40) showed low disease incidence even at higher blood pressure levels, while middle-aged and older groups demonstrated steeper increases. The largest sample sizes were concentrated in moderate blood pressure categories (BP=40-60).

<br>

### Income and Diet Quality
```{r}
color_palette <- colorRampPalette(c("lightblue", "darkblue"))(length(unique(ukb_data$diet_quality)))

figure_4 <- ukb_data %>%
  filter(!is.na(income) & !is.na(diet_quality)) %>%
  ggplot(aes(x = as.factor(income), fill = as.factor(diet_quality))) +
  geom_bar(position = "stack") +
  theme_minimal() +
  scale_fill_manual(values = color_palette) +
  labs(x = "Income Levels",
       y = "Count",
       fill = "Diet Quality",
       title = "Income vs Diet Quality") +
  theme(legend.position = "bottom")

ggplotly(figure_4)
```

Diet quality showed a positive correlation with income levels. Participants in the highest income bracket had the highest proportions of top diet quality, while those in the lowest income bracket exhibited the poorest diet quality. Middle-income groups displayed more balanced distributions.

<br>

### Sleep Duration Across Age Groups
```{r}
figure_5 <- ukb_data %>% 
  filter(!is.na(age) & !is.na(sleep_hour)) %>%
  mutate(age_group = cut(age, breaks = seq(10, 100, by = 10), right = FALSE, 
                         labels = paste(seq(10, 90, by = 10), "-", 
                                        seq(19, 99, by = 10), sep = ""))) %>%
  ggplot(aes(x = age_group, y = sleep_hour, fill = age_group)) +
  geom_boxplot() +
  theme_minimal() +
  labs(x = "Age Groups", 
       y = "Sleep Hours per Night",
       title = "Sleep Hours per Night by Age Groups",
       fill = "Age Groups") + 
  scale_fill_manual(values = c("lightgreen", "lightpink", "lightblue", 
                               "lightcoral", "lightgoldenrod"))

figure_5
```

Sleep duration varied slightly by age. Younger participants (30-39) had the shortest median sleep duration and the highest prevalence of short sleep (<6 hours). Older age groups (60-69, 70-79) displayed greater variability, with outliers on both ends of the sleep duration spectrum, suggesting more diverse sleep patterns in older adults.

<br> 

### Exposure variables and disease outcomes
```{r message = FALSE, warning = FALSE}
quantiles <- ukb_data %>%
  group_by(nafld_outcome) %>%
  filter(!is.na(b_lympho)) %>%
  summarize(
    q25 = quantile(b_lympho, 0.25),
    median = median(b_lympho),
    q75 = quantile(b_lympho, 0.75))

figure_14 <- ukb_data %>% 
  group_by(nafld_outcome) %>% 
  filter(!is.na(b_lympho)) %>% 
  mutate(
    lower_bound = quantile(b_lympho, 0.25) - 1.5 * IQR(b_lympho),
    upper_bound = quantile(b_lympho, 0.75) + 1.5 * IQR(b_lympho)
  ) %>%
  filter(b_lympho >= lower_bound & b_lympho <= upper_bound) %>% 
  ggplot(aes(x = b_lympho, fill = as.factor(nafld_outcome))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("#BCBDDC", "lightyellow"), name = "NAFLD Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Lymphocyte",
    y = "Density",
    title = "Density Plot of b_lympho by NAFLD Outcome"
  ) +
  theme_minimal(base_size = 14) + 
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  ) +  
  geom_vline(data = quantiles, aes(xintercept = q25, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = median, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "solid", size = 1, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = q75, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c("#8A2BE2", "#FDAE6B"))


quantiles <- ukb_data %>%
  group_by(nafld_outcome) %>%
  filter(!is.na(b_mono)) %>%
  summarize(
    q25 = quantile(b_mono, 0.25),
    median = median(b_mono),
    q75 = quantile(b_mono, 0.75)
  )

figure_15 <-ukb_data %>% 
  group_by(nafld_outcome) %>% 
  filter(!is.na(b_mono)) %>% 
  mutate(
    lower_bound = quantile(b_mono, 0.25) - 1.5 * IQR(b_mono),
    upper_bound = quantile(b_mono, 0.75) + 1.5 * IQR(b_mono)
  ) %>%
  filter(b_mono >= lower_bound & b_mono <= upper_bound) %>% 
  ggplot(aes(x = b_mono, fill = as.factor(nafld_outcome))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("#BCBDDC", "lightyellow"), name = "NAFLD Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Monocytes",
    y = "Density",
    title = "Density Plot of b_mono by NAFLD Outcome"
  ) +
  theme_minimal(base_size = 14) + 
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  ) +  
  geom_vline(data = quantiles, aes(xintercept = q25, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = median, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "solid", size = 1, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = q75, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c("#8A2BE2", "#FDAE6B"))

quantiles <- ukb_data %>%
  group_by(nafld_outcome) %>%
  filter(!is.na(b_neutro)) %>%
  summarize(
    q25 = quantile(b_neutro, 0.25),
    median = median(b_neutro),
    q75 = quantile(b_neutro, 0.75)
  )

figure_16 <- ukb_data %>% 
  group_by(nafld_outcome) %>% 
  filter(!is.na(b_neutro)) %>% 
  mutate(
    lower_bound = quantile(b_neutro, 0.25) - 1.5 * IQR(b_neutro),
    upper_bound = quantile(b_neutro, 0.75) + 1.5 * IQR(b_neutro)
  ) %>%
  filter(b_neutro >= lower_bound & b_neutro <= upper_bound) %>% 
  ggplot(aes(x = b_neutro, fill = as.factor(nafld_outcome))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("#BCBDDC", "lightyellow"), name = "NAFLD Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Neutrophils",
    y = "Density",
    title = "Density Plot of b_neutro by NAFLD Outcome"
  ) +
  theme_minimal(base_size = 14) + 
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  ) +  
  geom_vline(data = quantiles, aes(xintercept = q25, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = median, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "solid", size = 1, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = q75, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c("#8A2BE2", "#FDAE6B"))


quantiles <- ukb_data %>%
  group_by(nafld_outcome) %>%
  filter(!is.na(b_plate)) %>%
  summarize(
    q25 = quantile(b_plate, 0.25),
    median = median(b_plate),
    q75 = quantile(b_plate, 0.75)
  )

figure_17 <- ukb_data %>% 
  group_by(nafld_outcome) %>% 
  filter(!is.na(b_plate)) %>% 
  mutate(
    lower_bound = quantile(b_plate, 0.25) - 1.5 * IQR(b_plate),
    upper_bound = quantile(b_plate, 0.75) + 1.5 * IQR(b_plate)
  ) %>%
  filter(b_plate >= lower_bound & b_plate <= upper_bound) %>% 
  ggplot(aes(x = b_plate, fill = as.factor(nafld_outcome))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("#BCBDDC", "lightyellow"), name = "NAFLD Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Platelets",
    y = "Density",
    title = "Density Plot of b_plate by NAFLD Outcome"
  ) +
  theme_minimal(base_size = 14) + 
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  ) +  
  geom_vline(data = quantiles, aes(xintercept = q25, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = median, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "solid", size = 1, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = q75, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c("#8A2BE2", "#FDAE6B"))

grid.arrange(figure_14, figure_15, figure_16, figure_17, nrow = 2, ncol = 2)
```

For the continuous biomarkers (LMR, SII, and NPAR), values are consistently lower in the NAFLD group compared to the Healthy group, indicating a complex immune-inflammatory profile in NAFLD pathogenesis. Lower LMR reflects immune imbalance, potentially favoring innate over adaptive immunity, while decreased SII suggests a unique inflammatory response with possible compensatory mechanisms. Similarly, lower NPAR in NAFLD may indicate early changes in liver function and neutrophil activity. Despite these differences, the close median values and significant distribution overlap suggest limited discriminative power for these biomarkers alone in distinguishing NAFLD from healthy individuals. The NPS distribution shows higher proportions of healthy individuals across all scores, particularly in lower-risk categories (0-1), highlighting that while NAFLD alters immune-inflammatory profiles, this does not necessarily translate into higher prognostic risks.

<br>

```{r message = FALSE, warning = FALSE}
quantiles <- ukb_data %>%
  group_by(cirrho_outcome) %>%
  filter(!is.na(b_lympho)) %>%
  summarize(
    q25 = quantile(b_lympho, 0.25),
    median = median(b_lympho),
    q75 = quantile(b_lympho, 0.75))

figure_18 <- ukb_data %>% 
  group_by(cirrho_outcome) %>% 
  filter(!is.na(b_lympho)) %>% 
  mutate(
    lower_bound = quantile(b_lympho, 0.25) - 1.5 * IQR(b_lympho),
    upper_bound = quantile(b_lympho, 0.75) + 1.5 * IQR(b_lympho)
  ) %>%
  filter(b_lympho >= lower_bound & b_lympho <= upper_bound) %>% 
  ggplot(aes(x = b_lympho, fill = as.factor(cirrho_outcome))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("#BCBDDC", "lightyellow"), name = "Cirrhosis Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Lymphocyte",
    y = "Density",
    title = "Density Plot of b_lympho by Cirrhosis Outcome"
  ) +
  theme_minimal(base_size = 14) + 
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  ) +  
  geom_vline(data = quantiles, aes(xintercept = q25, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = median, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "solid", size = 1, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = q75, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c("#8A2BE2", "#FDAE6B"))


quantiles <- ukb_data %>%
  group_by(cirrho_outcome) %>%
  filter(!is.na(b_mono)) %>%
  summarize(
    q25 = quantile(b_mono, 0.25),
    median = median(b_mono),
    q75 = quantile(b_mono, 0.75)
  )

figure_19 <-ukb_data %>% 
  group_by(cirrho_outcome) %>% 
  filter(!is.na(b_mono)) %>% 
  mutate(
    lower_bound = quantile(b_mono, 0.25) - 1.5 * IQR(b_mono),
    upper_bound = quantile(b_mono, 0.75) + 1.5 * IQR(b_mono)
  ) %>%
  filter(b_mono >= lower_bound & b_mono <= upper_bound) %>% 
  ggplot(aes(x = b_mono, fill = as.factor(cirrho_outcome))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("#BCBDDC", "lightyellow"), name = "Cirrhosis Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Monocytes",
    y = "Density",
    title = "Density Plot of b_mono by Cirrhosis Outcome"
  ) +
  theme_minimal(base_size = 14) + 
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  ) +  
  geom_vline(data = quantiles, aes(xintercept = q25, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = median, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "solid", size = 1, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = q75, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c("#8A2BE2", "#FDAE6B"))

quantiles <- ukb_data %>%
  group_by(cirrho_outcome) %>%
  filter(!is.na(b_neutro)) %>%
  summarize(
    q25 = quantile(b_neutro, 0.25),
    median = median(b_neutro),
    q75 = quantile(b_neutro, 0.75)
  )

figure_20 <- ukb_data %>% 
  group_by(cirrho_outcome) %>% 
  filter(!is.na(b_neutro)) %>% 
  mutate(
    lower_bound = quantile(b_neutro, 0.25) - 1.5 * IQR(b_neutro),
    upper_bound = quantile(b_neutro, 0.75) + 1.5 * IQR(b_neutro)
  ) %>%
  filter(b_neutro >= lower_bound & b_neutro <= upper_bound) %>% 
  ggplot(aes(x = b_neutro, fill = as.factor(cirrho_outcome))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("#BCBDDC", "lightyellow"), name = "Cirrhosis Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Neutrophils",
    y = "Density",
    title = "Density Plot of b_neutro by Cirrhosis Outcome"
  ) +
  theme_minimal(base_size = 14) + 
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  ) +  
  geom_vline(data = quantiles, aes(xintercept = q25, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = median, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "solid", size = 1, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = q75, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c("#8A2BE2", "#FDAE6B"))


quantiles <- ukb_data %>%
  group_by(cirrho_outcome) %>%
  filter(!is.na(b_plate)) %>%
  summarize(
    q25 = quantile(b_plate, 0.25),
    median = median(b_plate),
    q75 = quantile(b_plate, 0.75)
  )

figure_21 <- ukb_data %>% 
  group_by(cirrho_outcome) %>% 
  filter(!is.na(b_plate)) %>% 
  mutate(
    lower_bound = quantile(b_plate, 0.25) - 1.5 * IQR(b_plate),
    upper_bound = quantile(b_plate, 0.75) + 1.5 * IQR(b_plate)
  ) %>%
  filter(b_plate >= lower_bound & b_plate <= upper_bound) %>% 
  ggplot(aes(x = b_plate, fill = as.factor(cirrho_outcome))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("#BCBDDC", "lightyellow"), name = "Cirrhosis Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Platelets",
    y = "Density",
    title = "Density Plot of b_plate by Cirrhosis Outcome"
  ) +
  theme_minimal(base_size = 14) + 
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  ) +  
  geom_vline(data = quantiles, aes(xintercept = q25, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = median, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "solid", size = 1, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = q75, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c("#8A2BE2", "#FDAE6B"))

grid.arrange(figure_18, figure_19, figure_20, figure_21, nrow = 2, ncol = 2)
```

For the Cirrhosis group, distinct differences are observed compared to the Healthy group: LMR and SII are lower in cirrhosis, reflecting immune dysregulation and potential immune exhaustion, while NPAR is elevated, indicating impaired liver function. This pattern suggests a shift from balanced immune and inflammatory responses in healthy individuals to a dysregulated state in cirrhosis. Despite these differences, the distribution overlap between the groups indicates some shared characteristics. The NPS distribution further highlights the contrast, with minimal representation of cirrhosis cases across risk categories, indicating that these immune-inflammatory changes do not strongly impact NPS scores in cirrhosis.

<br>

### Blood parameters and disease outcomes
```{r message = FALSE, warning = FALSE}
quantiles <- ukb_data %>%
  group_by(nafld_outcome) %>%
  filter(!is.na(expo_lmr)) %>%
  summarize(
    q25 = quantile(expo_lmr, 0.25),
    median = median(expo_lmr),
    q75 = quantile(expo_lmr, 0.75))

figure_6 <- ukb_data %>% 
  group_by(nafld_outcome) %>% 
  filter(!is.na(expo_lmr)) %>% 
  mutate(
    lower_bound = quantile(expo_lmr, 0.25) - 1.5 * IQR(expo_lmr),
    upper_bound = quantile(expo_lmr, 0.75) + 1.5 * IQR(expo_lmr)
  ) %>%
  filter(expo_lmr >= lower_bound & expo_lmr <= upper_bound) %>% 
  ggplot(aes(x = expo_lmr, fill = as.factor(nafld_outcome))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("skyblue", "pink"), name = "NAFLD Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Lymphocyte-to-Monocyte Ratio (LMR)",
    y = "Density",
    title = "Density Plot of expo_lmr by NAFLD Outcome"
  ) +
  theme_minimal(base_size = 14) + 
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  ) +  
  geom_vline(data = quantiles, aes(xintercept = q25, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = median, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "solid", size = 1, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = q75, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c("dodgerblue", "coral"))


quantiles <- ukb_data %>%
  group_by(nafld_outcome) %>%
  filter(!is.na(expo_sii)) %>%
  summarize(
    q25 = quantile(expo_sii, 0.25),
    median = median(expo_sii),
    q75 = quantile(expo_sii, 0.75)
  )

figure_7 <-ukb_data %>% 
  group_by(nafld_outcome) %>% 
  filter(!is.na(expo_sii)) %>% 
  mutate(
    lower_bound = quantile(expo_sii, 0.25) - 1.5 * IQR(expo_sii),
    upper_bound = quantile(expo_sii, 0.75) + 1.5 * IQR(expo_sii)
  ) %>%
  filter(expo_sii >= lower_bound & expo_sii <= upper_bound) %>% 
  ggplot(aes(x = expo_sii, fill = as.factor(nafld_outcome))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("skyblue", "pink"), name = "NAFLD Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Systemic Immune-Inflammation Index (SII)",
    y = "Density",
    title = "Density Plot of expo_sii by NAFLD Outcome"
  ) +
  theme_minimal(base_size = 14) + 
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  ) +  
  geom_vline(data = quantiles, aes(xintercept = q25, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = median, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "solid", size = 1, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = q75, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c("dodgerblue", "coral"))

quantiles <- ukb_data %>%
  group_by(nafld_outcome) %>%
  filter(!is.na(expo_npar)) %>%
  summarize(
    q25 = quantile(expo_npar, 0.25),
    median = median(expo_npar),
    q75 = quantile(expo_npar, 0.75)
  )

figure_8 <-ukb_data %>% 
  group_by(nafld_outcome) %>% 
  filter(!is.na(expo_npar)) %>% 
  mutate(
    lower_bound = quantile(expo_npar, 0.25) - 1.5 * IQR(expo_npar),
    upper_bound = quantile(expo_npar, 0.75) + 1.5 * IQR(expo_npar)
  ) %>%
  filter(expo_npar >= lower_bound & expo_npar <= upper_bound) %>% 
  ggplot(aes(x = expo_npar, fill = as.factor(nafld_outcome))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("skyblue", "pink"), name = "NAFLD Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Neutrophil-to-Albumin Ratio (NPAR)",
    y = "Density",
    title = "Density Plot of expo_npar by NAFLD Outcome"
  ) +
  theme_minimal(base_size = 14) + 
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  ) +  
  geom_vline(data = quantiles, aes(xintercept = q25, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = median, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "solid", size = 1, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = q75, 
                                   color = as.factor(nafld_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c("dodgerblue", "coral"))


figure_9 <-ukb_data %>% 
  filter(!is.na(expo_nps)) %>% 
  group_by(nafld_outcome, expo_nps) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  mutate(total_count = sum(count), 
         percentage = count / total_count * 100) %>%  
  ggplot(aes(x = as.factor(expo_nps), y = percentage, fill = as.factor(nafld_outcome))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), alpha = 0.8) +  
  geom_text(aes(label = paste0(round(percentage, 1), "%")),  
            position = position_dodge(width = 0.9), vjust = 0.75, size = 3) + 
  scale_fill_manual(values = c("skyblue", "pink"), name = "NAFLD Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Naples Prognostic Score (NPS)",
    y = "Percentage of Total Population",
    title = "Percentage of Total Population by NPS and NAFLD Outcome"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  ) 

grid.arrange(figure_6, figure_7, figure_8, figure_9, nrow = 2, ncol = 2)
```

For blood cell parameters (lymphocytes, monocytes, neutrophils, and platelets), the Healthy group consistently shows higher counts than the NAFLD group, reflecting preserved immune and hematopoietic function. Lower counts in NAFLD suggest a general depression of immune capacity and altered hematopoiesis associated with the disease. Despite this, the significant overlap in distributions suggests that individual blood cell counts alone may not serve as reliable diagnostic markers for NAFLD.

<br>

```{r message = FALSE, warning = FALSE}
quantiles <- ukb_data %>%
  group_by(cirrho_outcome) %>%
  filter(!is.na(expo_lmr)) %>%
  summarize(
    q25 = quantile(expo_lmr, 0.25),
    median = median(expo_lmr),
    q75 = quantile(expo_lmr, 0.75))

figure_10 <- ukb_data %>% 
  group_by(cirrho_outcome) %>% 
  filter(!is.na(expo_lmr)) %>% 
  mutate(
    lower_bound = quantile(expo_lmr, 0.25) - 1.5 * IQR(expo_lmr),
    upper_bound = quantile(expo_lmr, 0.75) + 1.5 * IQR(expo_lmr)
  ) %>%
  filter(expo_lmr >= lower_bound & expo_lmr <= upper_bound) %>% 
  ggplot(aes(x = expo_lmr, fill = as.factor(cirrho_outcome))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("skyblue", "pink"), name = "Cirrhosis Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Lymphocyte-to-Monocyte Ratio (LMR)",
    y = "Density",
    title = "Density Plot of expo_lmr by Cirrhosis Outcome"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  )  + 
  geom_vline(data = quantiles, aes(xintercept = q25, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = median, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "solid", size = 1, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = q75, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c("dodgerblue", "coral"))

quantiles <- ukb_data %>%
  group_by(cirrho_outcome) %>%
  filter(!is.na(expo_sii)) %>%
  summarize(
    q25 = quantile(expo_sii, 0.25),
    median = median(expo_sii),
    q75 = quantile(expo_sii, 0.75)
  )

figure_11 <- ukb_data %>% 
  group_by(cirrho_outcome) %>% 
  filter(!is.na(expo_sii)) %>% 
  mutate(
    lower_bound = quantile(expo_sii, 0.25) - 1.5 * IQR(expo_sii),
    upper_bound = quantile(expo_sii, 0.75) + 1.5 * IQR(expo_sii)
  ) %>%
  filter(expo_sii >= lower_bound & expo_sii <= upper_bound) %>% 
  ggplot(aes(x = expo_sii, fill = as.factor(cirrho_outcome))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("skyblue", "pink"), name = "Cirrhosis Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Systemic Immune-Inflammation Index (SII)",
    y = "Density",
    title = "Density Plot of expo_sii by Cirrhosis Outcome"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  )  + 
  geom_vline(data = quantiles, aes(xintercept = q25, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = median, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "solid", size = 1, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = q75, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c("dodgerblue", "coral"))

quantiles <- ukb_data %>%
  group_by(cirrho_outcome) %>%
  filter(!is.na(expo_npar)) %>%
  summarize(
    q25 = quantile(expo_npar, 0.25),
    median = median(expo_npar),
    q75 = quantile(expo_npar, 0.75)
  )

figure_12 <- ukb_data %>% 
  group_by(cirrho_outcome) %>% 
  filter(!is.na(expo_npar)) %>% 
  mutate(
    lower_bound = quantile(expo_npar, 0.25) - 1.5 * IQR(expo_npar),
    upper_bound = quantile(expo_npar, 0.75) + 1.5 * IQR(expo_npar)
  ) %>%
  filter(expo_npar >= lower_bound & expo_npar <= upper_bound) %>% 
  ggplot(aes(x = expo_npar, fill = as.factor(cirrho_outcome))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("skyblue", "pink"), name = "Cirrhosis Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Neutrophil-to-Albumin Ratio (NPAR)",
    y = "Density",
    title = "Density Plot of expo_npar by Cirrhosis Outcome"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  )  + 
  geom_vline(data = quantiles, aes(xintercept = q25, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = median, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "solid", size = 1, show.legend = FALSE) +
  geom_vline(data = quantiles, aes(xintercept = q75, 
                                   color = as.factor(cirrho_outcome)), 
             linetype = "dashed", size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c("dodgerblue", "coral"))

figure_13 <- ukb_data %>% 
  filter(!is.na(expo_nps)) %>% 
  group_by(cirrho_outcome, expo_nps) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  mutate(total_count = sum(count), 
         percentage = count / total_count * 100) %>%  
  ggplot(aes(x = as.factor(expo_nps), y = percentage, fill = as.factor(cirrho_outcome))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), alpha = 0.8) +  
  geom_text(aes(label = paste0(round(percentage, 1), "%")),  
            position = position_dodge(width = 0.9), vjust = 0.5, size = 3) + 
  scale_fill_manual(values = c("skyblue", "pink"), name = "Cirrhosis Outcome", 
                    labels = c("Healthy", "Disease")) +
  labs(
    x = "Naples Prognostic Score (NPS)",
    y = "Percentage of Total Population",
    title = "Percentage of Total Population by NPS and Cirrhosis Outcome"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",               
    plot.title = element_text(size = 8),      
    axis.title.x = element_text(size = 8),    
    axis.title.y = element_text(size = 8),   
    axis.text.x = element_text(size = 8),     
    axis.text.y = element_text(size = 8), 
    legend.title = element_text(size = 8),      
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  ) 

grid.arrange(figure_10, figure_11, figure_12, figure_13, nrow = 2, ncol = 2)
```

For these blood parameters, the Healthy group shows generally higher distribution values compared to the Cirrhosis group. Looking at the specific median lines, lymphocytes show almost identical median values between healthy and cirrhosis groups, while monocytes and neutrophils demonstrate higher median values in the cirrhosis group compared to healthy individuals, suggesting increased innate immune cell activation in liver disease. In contrast, platelets show a notably higher median in the healthy group compared to cirrhosis, with the most pronounced difference among all parameters, likely reflecting the impact of portal hypertension and possible splenic sequestration in cirrhosis. Despite the similar lymphocyte medians, the overall distribution reveals generally higher lymphocyte values in healthy individuals, indicating better preserved adaptive immunity. Distribution overlaps exist between groups across all parameters, though the platelet parameter shows the clearest separation between healthy and cirrhosis populations.
